---
title: "Homework 7"
author: "Ron Cordell, Lei Yang, Subhashini Raghunathan"
date: "March 23, 2016"
output: pdf_document
---
  

#Question 1:
**1.1 Load hw07_series1.csv**
**1.2 Describe the basic structure of the data and provide summary statistics of the series**

```{r}
setwd('C:/Subha/WS271-Regression/Labs/data')

#read the series from file
s1 = read.csv("hw07_series1.csv")

#describe the basic structure and summary stats
str(s1)
summary(s1)
hist(s1$X10.01, breaks="FD", main="Histogram of the series Homework7")
qqnorm(s1$X10.01,main="qqplot of the series Homework7", type ="p")
qqline(s1$X10.01)

```


- Basic structure - the series has 74 observations with no frequncy. The data are somewhat normally distributed with a positive skew.



**1.3 Plot histogram and time-series plot of the series. Describe the patterns exhibited in histograrm and**
**time-series plot. For time series analysis, is it sufficient to use only histogram to describe a series?**

```{r}
#plot time series
s1.ts= ts(data=s1$X10.01)
plot(s1.ts, main="time series plot of Homework7")
```

- Patterns - the time series plot has upwards and downwards trends that are somewhat persistent. It resembles a random walk model.

- There is no frequency attached to the time series, hence it is not possible to run decompose() and examine trends and seasonal variations.

- For time series analysis, histogram alone is not enough. ACF and PACF graphs much more important, as they show the dependency structures in the data.



**1.4 Plot the ACF and PACF of the series. Describe the patterns exhibited in the ACF and PACF.**

```{r}
acf(s1.ts, main="ACF of the series Homework7")
pacf(s1.ts, main="PACF of the series Homework7")
```

- The ACF is significant until lag 7. It gradually drops to 0.
- The PACF abruptly drops off to 0 after the second lag. This is indicative of an AR(2) model.



**1.5 Estimate the series using the ar() function.**
**1.6 Report the estimated AR parameters, the order of the model, and standard errors.**
```{r}
s1.arfit <- ar(s1.ts, method = "mle")
str(s1.arfit)
s1.arfit$order # order of the AR model with lowest AIC
s1.arfit$ar    # parameter estimate
s1.arfit$aic   # AICs of the fit models
sqrt(s1.arfit$asy.var) # asy. standard error
```

- The estimated model is AR(2):
**x[t] = 0.48x[t-1] + 0.3x[t-2] + w[t]**

- The model is order: 2 with a standard error of 0.11 for each parameter.



#Question 2:
 
**2.1 Simulate a time series of lenght 100 for the following model. Name the series x.**
**x[t] = 5/6 x[t???1] ??? 1/6 x[t???2] + w**

**2.2 Plot the correlogram and partial correlogram for the simulated series. Comments on the plots.**

```{r}
set.seed(898)
x = arima.sim(n = 100, list(ar = c(0.83, -0.17)))

plot(x, main="Simulated time series of an AR(2) model (0.83, -0.17)")
str(x)

acf(x, main="ACF of a simulated AR(2) model (0.83, -0.17)")
pacf(x, main = "PACF of a simulated AR(2) model (0.83, -0.17)")
```

- The ACF graph has 3 significant lags (the 4th one is marginally significant) and gradually drops off to zero. 
- The ACF has some periodicity but not the damped cosine function typical of an AR model. 
- The PACF graph has 1 significant lags and drops off significantly after that.  This is suggestive of an AR(1) model.But we know our true modell is AR(2). The difference could be due to sampling variations.


**2.3 Estimate an AR model for this simulated series. Report the estimated AR parameters, standard erros,
and the order of the AR model.**

```{r}
x.arfit <- ar(x, method = "mle")
str(x.arfit)
x.arfit$order # order of the AR model with lowest AIC
x.arfit$ar    # parameter estimate
x.arfit$aic   # AICs of the fit models
sqrt(x.arfit$asy.var) # asy. standard error
x.arfit$aic
```

- The ar() function estimated an AR(1) model but we can see that the AIC difference for order 2 is very small, suggesting that either AR(1) or AR(2) model would be a good fit.



2.4 Construct a 95% confidence intervals for the parameter estimates of the estimated model. Do the "true"
mode parameters fall within the confidence intervals? Explain the 95% confidence intervals in this context.

```{r}
x.arfit$ar + c(2,-2)*sqrt(x.arfit$asy.var)
```

- The confidence interval in this case includes the true population parameter alpha1 = 0.83. The model was estimated as AR(1), hence the second true population parameter was not estimated.
- The 95% confidence interval is the interval within which, for 95% of the samples, the true population parameters lie. The interpretation is the same in the context of this problem.


**2.5 Is the estimated model stationary or non-stationary?**

The estimated model is AR(1):
x[t] = 0.7x[t-1] + w
 
The characteristic equation is (1- 0.7*B) = 0, and the root of the is B = 1/0.7 = 1.4
As the root is > 1, the estimated model is stationary.


**2.6 Plot the correlogram of the residuals of the estimated model. Comment on the plot.**

```{r}
acf(x.arfit$resid[-(1:x.arfit$order)], main="ACF of residuals of the time estimated AR(1) model(0.7)")
```

- The ACF of the residuals resembles that of white noise. This indicates that the estimated AR(1) model is a good fit to the data.



